---
title: "ECON436 Lab 4"
author: "Abdulla Almansoori, Alexander Firth, Nathan Seto"
date: "`r Sys.Date()`"
output: pdf_document
editor_options: 
  markdown: 
    wrap: 72
---

Load required libraries (to read excel files, to manipulate dataframes, to plot time series using ts.plot, to test for stationarity)

```{r setup, echo = TRUE}

# clean existing variables
rm(list=ls())

library(knitr)
library(formatR)
library(tidyverse)
library(readxl)
library(graphics)
library(lubridate)
library(tseries)
library(urca) # For ca.jo
library(tsDyn) # For VECM
opts_chunk$set(include = TRUE, message = FALSE, warning = FALSE, tidy = TRUE)
```


A. Monthly sales taxes modeling
Input project data and create VECM data

```{r}
# Input the sheet "Monthly" of the file "FinalData.xlsx
FinalData <- read_excel("~/coursework/Data Raw/FinalData.xlsx", sheet = 1)

Manufacturing <- read_excel("~/coursework/Data Raw/Manufacturing.xlsx")
Food_accomidation <- read_excel("~/coursework/Data Raw/Food and accomidation.xlsx")
retailemployees <- read_excel("~/coursework/Data Raw/retail employees.xlsx")
Transportation <- read_excel("~/coursework/Data Raw/Transportation.xlsx")

TrafficVolume <- read_excel("~/coursework/Data Raw/TrafficVolume.xlsx") 
# Data from https://dtdapps.coloradodot.info/otis/TrafficData#ui/0/0/1/station/000127/criteria/27425//false/true/
TrafficVolume_clean <- TrafficVolume[TrafficVolume$Year >= 2012 & TrafficVolume$Year <= 2024, ]
TrafficVolume_clean <- TrafficVolume_clean[, -c(1, 15)] |> arrange(Year)

long_Trans <- Transportation |> 
  pivot_longer(cols = -Year, names_to = "Month", values_to = "transportation_employment") |> 
  mutate(Month = match(Month, month.abb), .before = Month)

long_Retail <- retailemployees |> 
  pivot_longer(cols = -Year, names_to = "Month", values_to = "RT_employment") |> 
  mutate(Month = match(Month, month.abb), .before = Month)

long_Food <- Food_accomidation |> 
  pivot_longer(cols = -Year, names_to = "Month", values_to = "AFS_employment") |> 
  mutate(Month = match(Month, month.abb), .before = Month)

long_Manf <- Manufacturing |> 
  pivot_longer(cols = -Year, names_to = "Month", values_to = "MAN_employment") |> 
  mutate(Month = match(Month, month.abb), .before = Month)

long_Traffic <- TrafficVolume_clean %>%
  pivot_longer(cols = -Year, names_to = "Month", values_to = "Value") %>%
  mutate(
    Month = match(Month, month.abb),
    Month = as.Date(paste(Year, Month, "01", sep = "-")),
  ) %>%
  select(Month, Year, Value,)
long_Traffic <- long_Traffic |> mutate(Date = Month, .before = Month)
long_Traffic <- long_Traffic |> mutate(Month = lubridate::month(long_Traffic$Date))


# The number of cointegrating vectors depends on number of distinct BLS categories that you think are determined by the economic conditions of Larimer county.
# These variables should go first in the dataframe df. They should be followed by the variables that are driven state or country-level conditions.
# Then add additional data you want to add (licenses and/or data you think is cointegrated with the sales tax data).
# Then add the sales tax data.
# Log all variables
df <- cbind(FinalData, 
            transportation_employment = long_Trans$transportation_employment, 
            MAN_employment= long_Manf$MAN_employment, 
            AFS_employment = long_Food$AFS_employment, 
            RT_employment = long_Retail$RT_employment,
            long_Traffic$Value)
df <- df |> mutate(month = month(df$Month), .before = year)
df <- df |> select(-Month)
new_df <- df %>% select("TP","GP","SP", "PSP","MAN_employment","AFS_employment","RT_employment","long_Traffic$Value" )
ldf <- log(new_df)
CombinedTS <- ts(as.matrix(ldf), start = c(2012, 1), frequency = 12)

```


#========================================================
# Cointegrating vectors (long run)
#========================================================

```{r}
# Conduct the Johansen procedure
Johansen <- ca.jo(ldf, ecdet="none", type="eigen", K=12) # K = number of periods in a year (4 if quarterly data, 12 if monthly data)

summary(Johansen)
```

1. Coefficients
```{r}
Beta <- round(coefB(Johansen, r=8), 3) # r = number of cointegrating vectors
Beta
```


2. Error correction term

For each cointegrating vector, 
- create the corresponding error correction term by multiplying the cointegrating vector with the time series in your model.
- test if the error correction term is stationary
- plot the error correction term

```{r}
ECT <- as.matrix(ldf) %*% Beta

ECT_df <- as.data.frame(ECT)
colnames(ECT_df) <- paste0("ECT_", 1:ncol(ECT_df))

stationarity_results <- sapply(ECT_df, function(x) adf.test(x)$p.value)

print(stationarity_results)

ECT_ts <- ts(ECT, start = c(2012, 1), frequency = 12) 

ts.plot(ECT_ts, col = rainbow(ncol(ECT_ts)), 
        main = "Error Correction Terms",
        ylab = "ECT Value", xlab = "Time")

```

#========================================================
# Short run adjustment coefficients
#========================================================

```{r}
vecm <- VECM(ldf, lag=11, r=7, estim="ML", include = "const") # lag = K-1
summary(vecm)
```

For each vector, we need at least one adjustment speed to have the opposite sign of the corresponding cointegrating vector coefficient and be statistically significant.

Vec 1: 
Vec 2:
....

